Exact Phylodynamics via Structured Markov Genealogy Processes

### 2023-07-20 ms.Rnw
some improvements

We use bold-face to mark random elements and the corresponding regular-face symbol to denote specific values of such elements.
Given any \cadlag\ process $f_t$, the set of jump-times is denoted $\event(f_t)=\{e\in[0,t]\;\vert\;\leftlim{f}_e\ne{f_e}\}$.

Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
We assume that $\N_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),
    \qquad
    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}

@@

Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\cadlag).
In fact, all of the processes we will describe in this paper will be taken to be \cadlag, and we will frequently need to refer to the left-limit.
Accordingly, if $\rdm{Z}_t$ is any \cadlag\ random process, we define
\begin{equation*}
  \leftlim{\rdm{Z}}_t\colonequals
  \begin{cases}
    \lim_{t'\uparrow{t}}\rdm{Z}_t, & t>0,\\
    \rdm{Z}_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\leftlim{\rdm{Z}}_t$ is thus left-continuous with right limits.

@@

Let us define the \emph{jump mark} process, $\U_t$, to be the mark of the latest jump as of time $t$.
As usual, we take the sample paths, $t\mapsto{\U_t(\omega)}$ for $\omega\in\Omega$, to be \cadlag.
Observe that $\U_t$ is a random, but not a Markov, process.

@@

For $\omega\in\Omega$, $t\mapsto(\X_t(\omega),\U_t(\omega))$ is a \cadlag\ function of time.
Let the \emph{history process}, $\Hist_t$, be the restriction of this random function to the interval $[0,t]$.
Note that $\Hist_t$ is a Markov process.

The non-explosiveness assumption implies that a.s., for every $t$, there is a finite, increasing sequence of random event times, which we denote by $\event(\Hist_t)\colonequals\left(\rdm{T}_k\right)_{k=1}^{\rdm{K}_t}$;
the length, $\rdm{K}_t$, of this sequence is itself random.
However, conditional on $\Hist_t$, $\rdm{T}_k$ and $\rdm{K}_t$, together with the mark process $\U_t$, and the population process $\X_t$ are deterministic.
One can write down an explicit probability measure for $\Hist_t$ in terms of these:
\begin{equation}\label{eq:Hdens}
  \begin{aligned}
    \pi^{\Hist}(\dd{\Hist_t})
    =&p_{0}(X_{0})\,\dd{X_0}\,\prod_{k=1}^{K_{t}}{\left(\alpha_{U_{T_k}}\left(T_k,\leftlim{X}_{T_k},X_{T_k}\right)\,\dd{X_{T_k}}\,\dd{T_k}\right)}\\
    &\qquad\times\exp{\left(-\sum_{u}{\int_{0}^{t}{\int{\alpha_u(t',\leftlim{X}_{t'},x')\,\dd{x'}}\,\dd{t'}}}\right)}.
  \end{aligned}
\end{equation}

@@

For our purposes, a \emph{genealogy} is a labeled, time-calibrated tree.
Its edges represent relationships of ancestry and descent among its nodes.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled extant lineages;
\item \emph{internal nodes}, which represent ancestral events;
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
More formally, we can take a genealogy, $G$, to be a finite sequence of internal nodes, together with a time.
Given $G$, the time is denoted $T(G)$; this is the time corresponding to the extant lineages.
The number of nodes is $K(G)$.
We order the nodes temporally, and denote the $k$-th node $\Node{p}_k(G)$ and we write $\Node{p}\in{G}$ if $\Node{p}$ is one of the nodes of $G$.
Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$, a parent, $\Anc(\Node{p})$, and a deme.
We let $\event(G)$ denote the sequence of node times:
$\event(G)=\left(\Node{p}_k(G)\right)_{k=1}^{K(G)}$, where $0=T(\Node{p}_1(G)){\le}T(\Node{p}_2(G)){\le}\dots{\le}T(\Node{p}_{K(G)}(G)){\le}T(G)$.
%% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
Root nodes are distinguished by being their own parents:
$\Node{p}$ is a root if and only if $\Anc(\Node{p})=\Node{p}$.
Every node also has one or more descendant nodes called \emph{children}.

@@

It will be useful to distill the elements that characterize a pruned genealogy.
Accordingly, given a pruned genealogy, $P$, we make the following definitions.
\begin{compactenum}[(a)]
\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $j\in\Zp$ and $t\in\Rp$, $Y^{P}_j(t)=\left(\ct^{P}_j(t),\anc^{P}_j(t),\deme^{P}_j(t)\right)$.
\end{compactenum}
For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
However, not every piecewise constant, \cadlag\ map $y:\Rp\to\Yspace$ defines a pruned genealogy.
If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.

Given $\eta\in\Yspace$, we will use the notation $\ct(\eta)$, $\anc(\eta)$, and $\deme(\eta)$ to refer to the coordinates of $\eta$.

@@

We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
Therefore, for $\eta,\eta'\in\Yspace$, and $i\in\Demes$, let us define
\begin{equation*}
  s_i(\eta,\eta')\colonequals{\left\vert\left\{j\in{\Zp}\;\vert\;\deme(\eta_j)=i\ \ \&\ \ \ct(\eta'_j)\ne\ct(\eta_j)\right\}\right\vert}, \qquad s(\eta,\eta')\colonequals(s_i(\eta,\eta'))_{i\in\Demes}\in{\Zp^\Demes}.
\end{equation*}
With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt(t),y(t))$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
In particular, $s(\eta,\eta)=0$ so that if $t\notin\event(P)$, then $s(\yt(t),y(t))=0$.
We refer to $s\left(\yt(t),y(t)\right)$ as the \emph{saturation} at time $t$.

@@

For $n,r,\ell,n\in{\Zp^\Demes}$, define the \emph{binomial ratio}
\begin{equation*}
  \BinRatio{n}{\ell}{r}{s}\colonequals
  \begin{cases}
    \frac{\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}{\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
    & \forall i\ n_i\ge{\{\ell_i,r_i\}}\ge{s_i}\ge{0}\\
    0, & \text{otherwise}
  \end{cases}
\end{equation*}
Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
Moreover, in consequence of the Chu-Vandermonde identity, we have
\begin{equation*}
  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
\end{equation*}
whenever $n_i\ge{\{\ell_i,r_i\}}\ge{0}$ for all $i$.

@@

Our first result will be an expression for the likelihood of a given pruned genealogy $P^*$ given the history $\Hist_t$ of the population process up to time $t=T(P^*)$.
Of course, not every pruned genealogy is compatible with $\Hist_t$.
In particular, $P^*$ is only compatible with $\Hist_t$ if every event in $P^*$ coincides with an event in $\Hist_t$.
That is, only if $\event(P^*)\subseteq{\event(\Hist_t)}$.
Moreover, certain events in $P^*$ may be incompatible with all possible histories.
For example, if $P^*$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*$ is incompatible with the population process itself.
Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(\eta,\eta')=1$ if and only if a change $\eta\to\eta'$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
More formally, $Q_u(\eta,\eta')=1$ if and only if there are $\omega\in\Omega$ and $t'<t\in\Rp$ such that $\U_{t'}(\omega)=u$, $\leftlim{Y}^{\prune(\Gen_t(\omega))}(t')=\eta$, and $Y^{\prune(\Gen_t(\omega))}(t')=\eta'$.
Recall that $n:\Xspace\to\Zp$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.

\begin{thm}\label{thm:pruned_lik}
  Let $P^*$ be a given pruned genealogy, $t=T(P^*)$, and $y^*=Y^{P^*}$.
  Then
  \begin{equation*}
    \begin{aligned}
      \Prob{P^*\vert\Hist_t}=&\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}\,Q_{U_e}(\yt^*(e),y^*(e))}\\
      &\qquad\times\Indicator{\event(P^*)\subseteq{\event(\Hist_t)}}.
    \end{aligned}
  \end{equation*}
\end{thm}
\begin{proof}
  As we have already observed, if $\event(P^*)\nsubseteq\event(\Hist_t)$, then $\Prob{P^*\vert\Hist_t}=0$.
  Similarly, if there is any event of $P^*$ which is incompatible with the population process, $\Prob{P^*\vert\Hist_t}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event(\Hist_t)$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, exactly $s_i$ of the drawn balls are black, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*$.
  Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*\vert\Hist_t}=\prod_{e\in\event(\Hist_t)}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
  \end{equation*}
  Returning to the possibility that $P^*$ is incompatible with $\Hist_t$, since $\Prob{P^*}=0$ if either any $Q_u=0$ or of any event of $P^*$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

@@

\begin{thm}
  Suppose we have a pruned genealogy $P^*_T$, $T>0$, with event times $\event(P^*_T)$.
  Let $y^*=Y^{\Prune^*_T}$.
  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \phi_u\big(x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\qquad
      \text{and} \qquad
      \beta_u(t,x,x')=\alpha_u(t,x,x')\,Q_u(y,y'),
    \end{gathered}
  \end{equation*}
  where $Q_u(y,y')$ is as in \cref{thm:pruned_lik}.
  Fix $\nu>0$ and suppose $w=w(t,x)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}=
      &\sum_u{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}
      -\sum_u{\int{w(t,x)\,\beta_u(t,x,x')\,\dd{x'}}}\\
      &\quad+\sum_{e\in\event(P^*_T)}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\frac{\beta_u(t,x',x)}{\nu}\,\phi_u\big(x,\yt^*_t,y^*_t\big)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x)=p_0(x)$.
  Then the likelihood of $P^*_T$ is
  \begin{equation*}
    \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
  \end{equation*}
\end{thm}

>>>>

Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\rdm{X}$), denotes a random element while the same symbol in regular face ($X$) denotes an element in its range.
We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$) and taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
We assume that $\N_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}

@@

Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\ie \cadlag).
In fact, all of the processes we will describe in this paper will be taken to have \cadlag\ sample paths, and we will frequently need to refer to their left-limits.
Accordingly, if $\rdm{Z}_t$ is any \cadlag\ random process, we define
\begin{equation*}
  \leftlim{\rdm{Z}}_t\colonequals
  \begin{cases}
    \lim_{t'\uparrow{t}}\rdm{Z}_t, & t>0,\\
    \rdm{Z}_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\leftlim{\rdm{Z}}_t$ is thus left-continuous with right limits.
Denote the set of event times of $\rdm{Z}$ by $\event{\rdm{Z}}{t}=\{e\in[0,t]\;\vert\;\leftlim{\rdm{Z}}_e\ne{\rdm{Z}_e}\}$.

@@

Let us define the \emph{jump mark} process, $\U_t$, to be the mark of the latest jump as of time $t$.
As usual, we take the sample paths, $t\mapsto{\U_t(\omega)}$ for $\omega\in\Omega$, to be \cadlag.
Observe that $(\X_t,\U_t)$ is a Markov process but that $\U_t$ is not.

@@

Consider the Markov process $(\X_t,\U_t)$.
We define its \emph{history process}, $\Hist_t$, to be the restriction of the random function $t\mapsto(\X_t,\U_t)$ to the interval $[0,t]$.
Note that $\Hist_t$ is itself a Markov process.
%% Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic, as of course are the Markov chains of event times $\rdm{\hat{T}}_k$ and the embedded chains $\rdm{\hat{X}}_k=\X_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\U_{\rdm{\hat{T}}_k}$.
Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic as are the embedded chain, $(\X_k,\U_k)$, and the point process of event times $\rdm{\hat{T}}_k$.
(Recall that $\rdm{\hat{X}}_k=\rdm{X}_{\rdm{\hat{T}}_k}$.)
The non-explosiveness assumption implies that, for every $t$, $\rdm{K}_t\coloneq\big\vert\event{\Hist}{t}\cap[0,t]\big\vert<\infty$ a.s.
The probability measure, $\pi^H$, for $\Hist_t$ can be expressed in terms of these:
\begin{equation*}%\label{eq:Hdens}
  \begin{aligned}
    \pi^{H}(\dd{H_t})
    =&p_{0}(\hat{X}_{0})\,\dd{\hat{X}_0}\,
    \prod_{k=1}^{K_{t}}{\left(\alpha_{\hat{U}_k}\left(\hat{T}_k,\hat{X}_{k-1},\hat{X}_{k}\right)\,\dd{\hat{X}_k}\,\dd{\hat{T}_k}\right)}\\
    &\qquad\times\exp{\left(-\sum_{k=1}^{K_t}{\int_{\hat{T}_{k-1}}^{\hat{T}_k}{\sum_{u}{\int{\alpha_{u}(t',\hat{X}_{k-1},x')\,\dd{x'}}}}\,\dd{t'}}\right)}.
  \end{aligned}
\end{equation*}

@@

For our purposes, a \emph{genealogy} is a labeled, time-calibrated tree.
Its edges represent relationships of ancestry and descent among its nodes.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled extant lineages;
\item \emph{internal nodes}, which represent ancestral events;
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
More formally, we can take a genealogy, $G$, to be a finite sequence of internal nodes, together with a time.
Given $G$, the time is denoted $T(G)$; this is the time corresponding to the extant lineages.
The number of nodes is $K(G)$.
We order the nodes temporally, and denote the $k$-th node $\Node{p}_k(G)$ and we write $\Node{p}\in{G}$ if $\Node{p}$ is one of the nodes of $G$.
Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$, a parent, $\Anc(\Node{p})$, and a deme.
Moreover, $\event{G}{t}$ is just the sequence of node times:
$\event{G}{t}=\left(T(\Node{p}_k(G))\right)_{k=1}^{K(G)}$.
%% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
Root nodes are distinguished by being their own parents:
$\Node{p}$ is a root if and only if $\Anc(\Node{p})=\Node{p}$.
Every node also has one or more descendant nodes called \emph{children}.

@@

It will be useful to distill the elements that characterize a pruned genealogy.
Accordingly, given a pruned genealogy, $P$, we make the following definitions.
\begin{compactenum}[(a)]
\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $t\in\Rp$,
  $Y^{P}(t)=\left(\ct^{P}(t),\anc^{P}(t),\deme^{P}(t)\right)$.
\end{compactenum}
For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
However, not every piecewise constant, \cadlag\ map $y:\Rp\to\Yspace$ defines a pruned genealogy.
If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.
Given $y\in\Yspace$, we will use the notation $\ct(y)$, $\anc(y)$, and $\deme(y)$ to refer to the coordinates of $y$.

@@

We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
Therefore, for $y,y'\in\Yspace$, and $i\in\Demes$, let us define
\begin{equation*}
  s_i(y,y')\colonequals{\left\vert\left\{j\in{\Zp}\;\vert\;\deme(y_j)=i\ \ \&\ \ \ct(y'_j)\ne\ct(y_j)\right\}\right\vert}, \qquad s(y,y')\colonequals(s_i(y,y'))_{i\in\Demes}\in{\Zp^\Demes}.
\end{equation*}
With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt(t),y(t))$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
In particular, $s(y,y)=0$ so that if $e\notin\event{P}{t}$, then $s(\yt(e),y(e))=0$.
We refer to $s\left(\yt(t),y(t)\right)$ as the \emph{saturation} at time $t$.

@@

For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
\begin{equation*}
  \BinRatio{n}{\ell}{r}{s}\colonequals
  \begin{cases}
    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
         & \forall i\ n_i\ge{\{\ell_i,r_i\}}\ge{s_i}\ge{0},\\
         0, & \text{otherwise}.
  \end{cases}
\end{equation*}
Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
Moreover, in consequence of the Chu-Vandermonde identity, we have
\begin{equation*}
  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
\end{equation*}
whenever $n_i\ge{\{\ell_i,r_i\}}\ge{0}$ for all $i$.

@@

Our first result will be an expression for the likelihood of a given pruned genealogy $P^*_t$ given the history $\Hist_t$ of the population process to time $t$.
Of course, not every pruned genealogy is compatible with $\Hist_t$.
In particular, $P^*_t$ is only compatible with $\Hist_t$ if every event in $P^*_t$ coincides with an event in $\Hist_t$.
That is, only if $\event{P^*}{t}\subseteq{\event{\Hist}{t}}$.
Moreover, certain events in $P^*_t$ may be incompatible with all possible histories.
For example, if $P^*_t$ has an event in which a lineage moves from deme $i$ to deme $i'$ but there are no $u\in\Jumps$ for which this is possible, then $P^*_t$ is incompatible with the population process itself.
Let us define the function $Q:\Jumps\times\Yspace\times\Yspace\to\{0,1\}$ so that $Q_u(y,y')=1$ if and only if a change $y\to{y'}$ in a pruned genealogy is compatible with an event of mark $u$ at that time.
Recall that $n:\Xspace\to\Zp^{\Demes}$ denotes the deme-occupancy function, $r^u$, the production of a jump of mark $u$, $\ell$, the lineage-count function, and $s$, the saturation.

\begin{thm}\label{thm:pruned_lik}
  Let $P^*_t$ be a given pruned genealogy and $y^*=Y^{P^*_t}$.
  Then
  \begin{equation*}
    \begin{aligned}
      \Prob{P^*_t\vert\Hist_t}=&\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}\,Q_{U_e}(\yt^*(e),y^*(e))}\\
      &\qquad\times\Indicator{\event{P^*}{t}\subseteq{\event{\Hist}{t}}}.
    \end{aligned}
  \end{equation*}
\end{thm}
\begin{proof}
  As we have already observed, if $\event{P^*}{t}\nsubseteq\event{\Hist}{t}$, then $\Prob{P^*_t\vert\Hist_t}=0$.
  Similarly, if there is any event of $P^*_t$ which is incompatible with the population process, $\Prob{P^*_t\vert\Hist_t}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event{\Hist}{t}$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*_t$, at time $e$, there are $\ell_i=\ell_i(y^*(t))$ lineages in deme $i$, of which $s_i=s_i(\yt^*(t),y^*(t))$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, exactly $s_i$ of the drawn balls are black, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*_t$.
  Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*_t\vert\Hist_t}=\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*(e))}{r^{U_e}}{s(\yt^*(e),y^*(e))}}.
  \end{equation*}
  Returning to the possibility that $P^*_t$ is incompatible with $\Hist_t$, since $\Prob{P^*_t}=0$ if either any $Q_u=0$ or any event of $P^*_t$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

@@

\begin{thm}
  Suppose that $P^*_T$ is a pruned genealogy, for $T>0$.
  Let $y^*=Y^{\Prune^*_T}$ and, for $x\in\Xspace$, $u\in\Jumps$, $t\le{T}$, define
  \begin{equation*}
    \phi_u\big(t,x)=\BinRatio{n(x)}{\ell(y^*_t)}{r^u}{s(\yt^*_t,y^*_t)}\,Q_u(\yt^*_t,y^*_t),
  \end{equation*}
  where $Q_u$ is as in \cref{thm:pruned_lik}.
  Suppose $w=w(t,x)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}=
      &\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}
      -\sum_u{\int{w(t,x)\,\alpha_u(t,x',x)\,\dd{x'}}}\\
      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_u{\int{w(t,x')\,\alpha_u(t,x',x)\,\phi_u(t,x)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x)=p_0(x)$.
  Then the likelihood of $P^*_T$ is
  \begin{equation*}
    \lik(P^*_T)=\int{w(T,x)\,\dd{x}}.
  \end{equation*}
\end{thm}

The final result shows how to compute the likelihood of an obscured genealogy.
